---
title: "Class 17: Obtaining and processing SRA datasets on AWS"
author: "Amy Nguyen: (PID: A18148284)"
format: pdf
---

## Core Unix Commands

Most Unix commands have super short names, which makes them quick to type but annoying to learn. Major file system related commands include:

cd: Change directory

pwd: Print working directory

ls: List files and directories

mkdir: Make a new directory

rm: Delete files and directories

cp: Copy files or dirs (source \> destination)

mv: Move files or dirs (basically re-name)

nano: A very basic text editor that is always available

less: To view/read text files page by page (paper program)

## AWS EC2 Instance

My AWS instance: ssh -i "bioinf_amynguyen.pem" ubuntu\@ec2-54-187-195-168.us-west-2.compute.amazonaws.com

To copy from my AWS instance: scp -i \~/Downloads/bioinf_amynguyen.pem ubuntu\@ec2-54-187-195-168.us-west-2.compute.amazonaws.com:\~/work/results.tsv .

## Class 17 AWS instance address

ssh -i "\~/Downloads/bioinf_amynguyen.pem" ubuntu\@ec2-35-94-26-238.us-west-2.compute.amazonaws.com

tar -zxvf kallisto_linux-v0.44.0.tar.gz

scp -r -i "\~/Downloads/bioinf_amynguyen.pem" ubuntu\@ec2-35-94-26-238.us-west-2.compute.amazonaws.com:\~/\*\_quant .

## Downstream analysis

```{r}
library(tximport)

# setup the folder and filenames to read
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)

```

```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## Principal Component Analysis

```{r}
pca <- prcomp(t(x), scale=TRUE)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

x

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()
```

## OPTIONAL: Differential-expression analysis

```{r}
library(DESeq2)

sampleTable <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(sampleTable) <- colnames(txi.kallisto$counts)

dds <- DESeqDataSetFromTximport(txi.kallisto,
                                sampleTable, 
                                ~condition)
```

```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)

```
