---
title: "Class 06 Function Homework"
author: "Amy Nguyen (PID: A18148284)"
format: pdf
---

# What is the function input? The function input is a character string that specifies the PDB ID of a specific protein structure.

# What does the function do and how do we use it? The function reads a protein structure, trims the structure to chain A and alpha carbon atoms, extracts the B-factor values, and plots the B-factor along the protein sequence. It can be used to visualize B-factor variation along the protein sequence

# What is the function output? The function output is a line plot of B-factor values for alpha carbon atons in chain A of the protein.

```{r}
library(bio3d)
```

```{r}

# Name the function
protein_analysis <- function(pdb_id) {

  # Read the protein structure
  pdb <- read.pdb(pdb_id)
  
  # Trim to specified chain A and alpha carbon atom 
  pdb_chain <- trim.pdb(pdb, chain = "A", elety = "CA")
  
  # Extract B-factor
  b_factors <- pdb_chain$atom$b 
  
  # Plot the B-factor
  plotb3(b_factors, sse=pdb_chain, typ="l", ylab="Bfactor")
  
}
```

```{r}
protein_analysis("4AKE")
```

