---
title: "Class 12 HW"
author: "Amy Nguyen (PID: A18148284)"
format: pdf
---

> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.
Sample size for each genotype: A/A - 108, A/G - 233, G/G - 121
Corresponding median expression levels: A/A - 31.25, A/G - 233, G/G - 121

How many samples do we have?

```{r}
# Import data from file 
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
# Find the sample size of each genotype 
table(expr$geno)
```

```{r}
# Looking at the five-number summary for each genotype
summary(expr$exp[expr$geno == "A/A"])
summary(expr$exp[expr$geno == "A/G"])
summary(expr$exp[expr$geno == "G/G"])
```
```{r}
# Find the median of each genotype
geno_boxplot <- boxplot(expr$exp ~ expr$geno, plot = FALSE)
geno_boxplot$stats[3, ]
```

Lets make a boxplot

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? The boxplot shows that those with the A/A genotype have higher median ORMDL3 expression compared to those with the G/G genotype, while those with A/G are in the middle. This suggests that the SNP is associated with ORMDL3 expression levels, where the A allele increases expression. 

Lets make a boxplot

```{r}
# Load ggplot
library(ggplot2)
```

```{r}
# Create a box plot from the given data 
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

